---
title: "fastQC reports"
author: "Sébastien Renaut (sebastien.renaut@criucpq.ulaval.ca)"
date: "`r Sys.time()`"
output:
  html_document:
    number_sections: T
params:
  fq.dir: 'data/fastq'
  qc.dir: 'fastqc_results'
  threads: 12
  plots_arg: FALSE
  fastqc.path: 'fastqc' 
  metadata: 'data/librairies_1_a_70_et_RIN.xlsx'
  adapters.dir: NULL
---


# Introduction  
* This is a fastQC report of raw sequences. I uses functions from `fastqcr`, which is itself a wrapper for [`fastqc`](https://www.bioinformatics.babraham.ac.uk/projects/fastqc/)
* You can knit this directly from `Rstudio` or you can create this document by running `Rscript -e "rmarkdown::render('fastqc_reports.Rmd',render = list())"` on a Linux/Unix machine. 

 
```{r, timehook,echo=F}
knitr::knit_hooks$set(timehook = local({
  now <- NULL
  function(before, options) {
    if (before) {
      # record the current time before each chunk
      now <<- Sys.time()
    } else {
      # calculate the time difference after a chunk
      res <- difftime(Sys.time(), now)
      # return a character string to show the time
      paste("\n\n\nTime for this code chunk to run:",signif(res,3),attr(res,'units'))
    }
  }
}))

knitr::opts_knit$set(root.dir = '..')
``` 

```{r, requirements,echo=TRUE,message =FALSE,warning =F}
library(fastqcr)
library(DT)
library(readxl)
library(dplyr)
library(ggplot2)
library(tidyr)
library(plotly)
library(patchwork)
source('R/fastqc_functions.R')
```

# FastQC parameters
```{r, parameters,message = T, echo =T}
print(params)
```

# Running FastQC functions
```{r, tables,message = T, eval = T, echo = T, timehook = T}
fastqc_statistics = fastqc_wrapper(fq.dir = params$fq.dir,
                qc.dir = params$qc.dir,
                threads = params$threads,
                fastqc.path =params$fastqc.path,
                metadata.dir = params$metadata,
                adapters.dir = params$adapters.dir)
```

# Summary table
  * Listing statistics for each .fastq file
```{r, summary_tables,echo = F}
  
  #datatable 
  datatable(fastqc_statistics[[1]],caption = htmltools::tags$caption(htmltools::strong("Summary statistics"), style="color:darkred"))
``` 


# Summary table (detailed)
  * Listing summary statistics for the whole cohort
```{r, summary_detailed,echo = F}
  #datatable
  datatable(fastqc_statistics[[2]],caption = htmltools::tags$caption(htmltools::strong("Summary statistics (detailed)"), style="color:darkred"))
```

# Summary Plots 
```{r, summary,echo =F,fig.show="hold",out.width="50%"}
fastqc_statistics[[1]]$type = ifelse(fastqc_statistics[[1]]$type=='Sain','Normal','Tumor')

rin_phred = ggplot(fastqc_statistics[[1]], aes(x=mean_quality, y=RIN,col = type)) + 
    geom_point()+
    xlab("Mean quality per sample (Phred Score)") +
    ylab("RNA Integrity Number") 

rin_nbreads = ggplot(fastqc_statistics[[1]], aes(x=tot.seq, y=RIN,col = type)) + 
    geom_point( show.legend = FALSE)+
    xlab("Total Nb of Sequences (10e6)") +
    ylab("RNA Integrity Number") 

hist_phred = ggplot(fastqc_statistics[[1]], aes(x=mean_quality)) +
    geom_histogram() + 
    xlab("Mean quality per sample (Phred Score)")

hist_reads = ggplot(fastqc_statistics[[1]], aes(x=tot.seq)) +
    geom_histogram() + 
    xlab("Total Nb of Sequences (10e6)")

hist_reads

hist_phred

if(params$plots_arg ==T){
ggplotly(rin_nbreads)

ggplotly(rin_phred)
}


pdf(file.path(params$qc.dir,'qc_plots.pdf'),width = 9,height = 7)
hist_reads + hist_phred


write.table(fastqc_statistics[[1]],file.path(params$qc.dir,'fastqc_statistics.csv'))



```


# QC Plots
 * A few plots that might be worth looking at (but remember that we have all individual plots).
```{r, plots,timehook =T,fig.width = 11, echo=F}
# Plot per sequence GC content
qc_read_collection = fastqc_statistics[[3]]
if(params$plots_arg ==T){

ggplotly(qc_plot_collection(qc_read_collection, "Per sequence GC content"))

# Per base sequence quality
ggplotly(qc_plot_collection(qc_read_collection, "Per base sequence quality"))

# Per sequence quality scores
ggplotly(qc_plot_collection(qc_read_collection, "Per sequence quality scores"))

# Overrepresented sequences
ggplotly(qc_plot_collection(qc_read_collection, "Overrepresented sequences"))

# Adapter content
if(is.numeric(qc_read_collection$adapter_content$Position)==F) {
newmean_position = c(1:length(qc_read_collection$adapter_content$Position))[regexpr('-',qc_read_collection$adapter_content$Position,fixed = T)>0]
means = strsplit(qc_read_collection$adapter_content$Position[newmean_position],'-')
means = sapply(means,function(x) mean(as.numeric(x)))
qc_read_collection$adapter_content$new_position = ''
qc_read_collection$adapter_content$new_position[newmean_position] = means
qc_read_collection$adapter_content$new_position[-newmean_position] = qc_read_collection$adapter_content$Position[-newmean_position]
qc_read_collection$adapter_content$new_position = as.numeric(qc_read_collection$adapter_content$new_position)
}
if(is.numeric(qc_read_collection$adapter_content$Position)==T) {
  qc_read_collection$adapter_content$new_position = qc_read_collection$adapter_content$Position
}

adapter_data = qc_read_collection$adapter_content[,-2] %>% pivot_longer(cols = !c(sample,new_position))

adapter_plot = adapter_data %>%
  ggplot( aes(x=new_position, y=value, group=sample, color=sample)) +
    geom_line() +
    ylim(0, 100) + ylab('percentage') + ggtitle('adapter (all) content')

ggplotly(adapter_plot)
}
```