---
title: "fastQC reports"
author: "Sébastien Renaut (sebastien.renaut@criucpq.ulaval.ca)"
date: "`r Sys.time()`"
output:
  html_document:
    number_sections: T
params:
  fq.dir: 'C:/Users/renseb01/Documents/rnaseq/data/fastq'
  qc.dir: 'C:/Users/renseb01/Documents/rnaseq/data/fastqc_results'
  threads: 12
  fastqc.path: 'C:/Program\ Files/FastQC/fastqc' 
---


# Introduction  
* This is a fastQC report of raw sequences. I uses functions from `fastqcr`, which is itself a wrapper for [`fastqc`](https://www.bioinformatics.babraham.ac.uk/projects/fastqc/)
* You can knit this directly from `Rstudio` or you can create this document by running `Rscript -e "rmarkdown::render('fastqc_reports.Rmd',params = list())"` on a Linux/Unix machine. 


 
 
```{r, timehook,echo=F}
knitr::knit_hooks$set(timehook = local({
  now <- NULL
  function(before, options) {
    if (before) {
      # record the current time before each chunk
      now <<- Sys.time()
    } else {
      # calculate the time difference after a chunk
      res <- difftime(Sys.time(), now)
      # return a character string to show the time
      paste("\n\n\nTime for this code chunk to run:",signif(res,3),attr(res,'units'))
    }
  }
}))
``` 

```{r, requirements,echo=TRUE}
library(fastqcr)
library(DT)
```

```{r, functions,echo = F}
###hack for windows
fastqc_windows = function (fq.dir = getwd(), qc.dir = NULL, threads = 8, fastqc.path = "") 
{
  fq.dir = sub('C:','/mnt/c',fq.dir)
  qc.dir = sub('C:','/mnt/c',qc.dir)
  fastqc.path = sub('C:','wsl.exe /mnt/c',fastqc.path)
  
  if (is.null(qc.dir)) 
    qc.dir <- file.path(fq.dir, "FASTQC")
  #.create_dir(qc.dir)
  cmd <- paste0(fastqc.path, " ", fq.dir, "/*  --threads ", 
                threads, " --outdir ", qc.dir)
  system(cmd)
}



fastqcr_wrapper = function(fq.dir = 'path',
                           qc.dir = 'path', 
                           threads = 8,
                           fastqc.path = 'path'){
  
  if(Sys.info()['sysname'] != 'Windows') {
    fastqc(qc.dir = qc.dir ,fq.dir= fq.dir,threads= threads,fastqc.path= fastqc.path)}
  
  if(Sys.info()['sysname'] == 'Windows') {
    message('waning: Running a modified version of fastqc() for Windows');
    fastqc_windows(qc.dir = qc.dir,
                   fq.dir = fq.dir,
                   threads = threads,
                   fastqc.path = fastqc.path)}
  
  
  #zip the whole thing
  zip = paste0('zip ',file.path(fq.dir,"../fastqc_individual_reports.zip "),file.path(qc.dir,"*fastqc.zip"))
  system(zip)
  
  message(paste0(Sys.time(),
                 '--- Zip archive is stored here: ',
                 file.path(fq.dir,"../fastqc_individual_reports.zip")))

  #aggregate results
  qc = suppressMessages(qc_aggregate(qc.dir = qc.dir, progressbar = F))
  
  # Generates a summary of qc_aggregate
  qc_summary = summary(qc)
  
  # General statistics of fastqc reports.
  qc_stats = qc_stats(qc)

  return(list(qc_summary,qc_stats))
}
```

# FastQC parameters
```{r, parameters,message = T, echo =T}
print(params)
```

# Running FastQC functions
```{r, tables,message = T, eval = T, echo = T, timehook = T}
fastqc_statitics= fastqcr_wrapper(fq.dir = params$fq.dir,
                qc.dir = params$qc.dir,
                threads = params$threads,
                fastqc.path = params$fastqc.path)
```

# Summary table
  * Listing statistics for each .fastq file
```{r, summary_tables,echo = F}
  #datatable 
  datatable(fastqc_statitics[[2]],caption = htmltools::tags$caption(htmltools::strong("Summary statistics"), style="color:darkred"))
``` 


# Summary table (detailed)
  * Listing summary statistics for the whole cohort
```{r, summary_detailed,echo = F}
  #datatable
  datatable(fastqc_statitics[[1]],caption = htmltools::tags$caption(htmltools::strong("Summary statistics (detailed)"), style="color:darkred"))
```


# QC Plots
 * A few plots that might be worth looking at (but remember that we have all individual plots).
```{r, plots,timehook =T, echo=F}
qc.files <- list.files(params$qc.dir,pattern = '.zip', full.names = TRUE)
qc.names = gsub("_fastqc.zip","",list.files(params$qc.dir,pattern = '.zip', full.names = F))
nb_samples <- length(qc.files)
modules = c("all")

qc_read_collection <- suppressMessages(qc_read_collection(qc.files, sample_names = qc.names,modules = modules))

# Plot per sequence GC content
qc_plot_collection(qc_read_collection, "Per sequence GC content")

# Per base sequence quality
qc_plot_collection(qc_read_collection, "Per base sequence quality")

# Per sequence quality scores
qc_plot_collection(qc_read_collection, "Per sequence quality scores")
```