---
title: "QC summary LORD"
author: "Sébastien Renaut (sebastien.renaut@criucpq.ulaval.ca)"
date: "`r Sys.Date()`"
output:
  html_document:
    number_sections: F
params:
  datapath: 'C:/Users/renseb01/Documents/rnaseq'
  outputpath: '../..' 
---



```{r setup, include=FALSE}
knitr::opts_knit$set(root.dir = params$datapath)
knitr::opts_chunk$set(echo = F)
library(DT) #data viz
library(ggplot2)
library(patchwork)
library(readxl)
library(dplyr)
```

# Data
* 736 samples analysed: 
  * **Part1: 64 samples**
  * **Part2: 416 samples**
  * **Part3: 256 samples**
  * **Part4: 288 samples**
* Alignments statistics based on a pseudo-random set of 200,000 sequences for each sample.
  * Percentage aligned correspond to the fraction of reads alignment to a transcribed region of the genome.
* **No samples are both below 70M reads AND below 70% aligned**.

```{r data}
part1  = read.delim('lordQC/fastqc_statistics_part1.csv',header = T,sep = ' ')
part2  = read.delim('lordQC/fastqc_statistics_part2.csv',header = T,sep = ' ')
part3  = read.delim('lordQC/fastqc_statistics_part3.csv',header = T,sep = ' ')
part4  = read.delim('lordQC/fastqc_statistics_part4.csv',header = T,sep = ' ')

#rbind
part1 = data.frame(part1,part = 'part1')
part2 = data.frame(part2,part = 'part2')
part3 = data.frame(part3,part = 'part3')
part4 = data.frame(part4,part = 'part4')
sequences = rbind(part1, part2, part3, part4)


#order and keep only one...
sequences = sequences[order(sequences$sample),]

#clean up columns
sequences = sequences[seq(1,nrow(sequences),by = 2),]
sequences  = sequences[,c(1,3,4,5,6,8,9,10,12)]
colnames(sequences)[4]  = 'total.seq(10e6)' 
sequences[,4] = round(sequences[,4],2)
```

***  


## Outliers (Alignments) 
```{r mini_alignments} 
part1 = read.csv('lordmini_alignments/expression_df_part1.csv', header = T)
part2 = read.csv('lordmini_alignments/expression_df_part2.csv', header = T)
part3 = read.csv('lordmini_alignments/expression_df_part3.csv', header = T)
part4 = read.csv('lordmini_alignments/expression_df_part4.csv', header = T)

align  = data.frame(reads = c(colSums(part1[-c(1:4),-1]),colSums(part2[-c(1:4),-1]),colSums(part3[-c(1:4),-1]),colSums(part4[-c(1:4),-1])))
align$sample = c(colnames(part1[,-1]),colnames(part2[,-1]),colnames(part3[,-1]),colnames(part4[,-1]))
align$bio_sample  = sapply( strsplit(align$sample,split="_i5.",fixed = T), "[", 2) 
align$percentage_aligned = round(align$reads/2000,2)
align = align[,c(3,1,4)]

align_sequences = merge(align,sequences,by = 'bio_sample')
too_low = align_sequences[align_sequences$percentage_aligned<70,]
too_low = too_low[order(too_low$percentage_aligned),]

#rownames(align) = NULL
#too_low = align[align$percentage_aligned<70,]
#too_low = too_low[order(too_low$percentage_aligned),]

datatable(too_low,caption = htmltools::tags$caption(htmltools::strong("ALIGNMENT PERCENTAGE BELOW 70%"), style="color:darkred"))
```



## Outliers (sequences)
```{r too many}
too_many_seq = align_sequences[align_sequences$`total.seq(10e6)` > 200,]
too_few_seq = align_sequences[align_sequences$`total.seq(10e6)` < 50,]
too = rbind(too_many_seq,too_few_seq)
datatable(too[order(too$`total.seq(10e6)`),],caption = htmltools::tags$caption(htmltools::strong("TOO MANY/FEW (below 50M) SEQUENCES"), style="color:darkred"))


datatable(too_low,caption = htmltools::tags$caption(htmltools::strong("ALIGNMENT PERCENTAGE BELOW 70%"), style="color:darkred"))
```



***  

# Summary plots 
```{r plots, fig.height = 8,message=FALSE}
align_sequences = merge(align,sequences,by = 'bio_sample')
align_sequences = align_sequences[align_sequences$`total.seq(10e6)`<200,]

#histograms
hist_reads = ggplot(align_sequences, aes(x=`total.seq(10e6)`)) +
    geom_histogram() + 
    xlab("Total Nb of Sequences (10e6)")+
   ggtitle("Sequencing outputs (992 Paired-Ends samples)")

hist_phred = ggplot(align_sequences, aes(x=mean_quality)) +
    geom_histogram() + 
    xlab("Phred Score") + 
     ggtitle("Mean quality of reads")
    

hist_align = ggplot(align_sequences, aes(x=percentage_aligned)) +
    geom_histogram() + 
    xlab("Percentage")+
    ggtitle("Percentage of Sequences aligned to transcribed\nregion of the genome (based on 200,000 reads)") 
  


#seqrin
align_sequences_rin = align_sequences[align_sequences$RIN>1,]
plot_seq_rin = ggplot(align_sequences_rin, aes(x=RIN, y=`total.seq(10e6)`)) +
  geom_point(position="dodge") +
  geom_smooth(method = "lm",formula = y ~ x) +
 ggtitle("Total number of sequences\nas a function of RIN score")

#seqtype
plot_seq_type = ggplot(align_sequences, aes(x=type, y=`total.seq(10e6)`, fill=type)) +
  geom_violin(position="dodge") +
  geom_jitter(shape=16, position=position_jitter(0.1)) +
  ggtitle("Number of sequences as a function of tissue type") +
  theme(legend.position = 'none')

#seqqual
plot_seq_qual = ggplot(align_sequences_rin, aes(x=mean_quality, y=`total.seq(10e6)`)) +
  geom_point(position="dodge") +
  geom_smooth(method = "lm",formula = y ~ x) +
 ggtitle("Number of sequences as a function of quality") +
  xlab("Phred Score")

#qualtissue
plot_qual_type = ggplot(align_sequences, aes(x=type, y=mean_quality, fill=type)) +
  geom_violin(position="dodge") + 
  geom_jitter(shape=16, position=position_jitter(0.1)) +
  ggtitle("Reads quality as a function of tissue type")+
  theme(legend.position = 'none')+
  ylab("Phred Score")
  

#
plot_seq_mean = ggplot(align_sequences_rin, aes(x=mean_quality, y=percentage_aligned)) +
  geom_point(position="dodge") +
  geom_smooth(method = "lm",formula = y ~ x) +
 ggtitle("Percentage of sequences aligned\nas a function of reads quality") +
  xlab("Phred Score")

#
plot_align_type = ggplot(align_sequences, aes(x=type, y=percentage_aligned, fill=type)) +
  geom_violin(position="dodge") + 
  geom_jitter(shape=16, position=position_jitter(0.1)) +
  ggtitle("Percentage of sequences aligned\nas a function of tissue type")+
  theme(legend.position = 'none')

#output
pdf('lordQC/LORD_qc.pdf',width = 15,height = 12) 
((hist_reads|hist_phred|hist_align) / (plot_seq_qual|plot_seq_rin|plot_seq_mean) / (plot_seq_type|plot_qual_type|plot_align_type)) +plot_annotation(tag_levels = list('1'))
dev.off()

#output
pdf('lordQC/violon.pdf',width = 15,height = 10) 
(plot_seq_type+plot_qual_type+plot_align_type) +plot_annotation(tag_levels = list('1'))
dev.off()

```



#dominique veut les stats pour une coupe d'invividus:
```{r correls}
resume = read_xlsx('2023-10-03_Resume_nombre_reads_RNAseq_500_ADK.xlsx',sheet = 'List_<70M_reads')

resume2 = merge(resume,align_sequences, by.x = 'Nom de librairie', by.y = 'bio_sample',all.x =T)

write.table(resume2, 'rnaseq/resume2.csv')

```





